---
title: "Brickyard 400 — Green-Flag Passing Analysis"
subtitle: "2026 NASCAR Cup Series · Indianapolis Motor Speedway"
output:
  html_document:
    theme: null
    css: gts_theme.css
    self_contained: true
    df_print: kable
params:
  data_file: "IndyPasses.xlsx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 5.2, dpi = 130,
  fig.align = "center", dev = "png"
)

library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(scales)
library(forcats)

# ---- GTS palette -----------------------------------------------------------
gts_gold   <- "#FFE500"
gts_dark   <- "#1a1a1a"
gts_panel  <- "#242424"
gts_grid   <- "#3a3a3a"
gts_text   <- "#f2f2f2"
gts_green  <- "#4CAF50"
gts_amber  <- "#FF9800"
gts_red    <- "#E53935"

theme_gts <- function(base = 13) {
  theme_minimal(base_size = base, base_family = "Helvetica") +
    theme(
      plot.background   = element_rect(fill = gts_dark,  color = NA),
      panel.background  = element_rect(fill = gts_panel, color = NA),
      panel.grid.major  = element_line(color = gts_grid, linewidth = 0.3),
      panel.grid.minor  = element_blank(),
      text              = element_text(color = gts_text),
      axis.text         = element_text(color = gts_text),
      axis.title        = element_text(color = gts_text, face = "bold"),
      plot.title        = element_text(color = gts_gold, face = "bold", size = base + 3),
      plot.subtitle     = element_text(color = "#bdbdbd", size = base - 1),
      legend.position   = "top",
      legend.title      = element_blank(),
      legend.key        = element_rect(fill = gts_panel, color = NA),
      plot.margin       = margin(12, 16, 12, 12)
    )
}

# ---- load ------------------------------------------------------------------
passes    <- read_xlsx(params$data_file, sheet = "Passes")
lapstatus <- read_xlsx(params$data_file, sheet = "LapStatus")
pitlaps   <- read_xlsx(params$data_file, sheet = "PitLaps")

# The NASCAR feed appends "(i)"/"#" to ineligible / rookie drivers. Cosmetic —
# strip so names read clean.
clean_driver <- function(x) str_trim(str_remove(x, "\\s*[\\(#].*$"))
passes  <- passes  %>% mutate(passer = clean_driver(passer),
                              passed = clean_driver(passed))
pitlaps <- pitlaps %>% mutate(driver = clean_driver(driver))

# true green-flag passes only
tg <- passes %>% filter(PassCategory == "True Green (8+)")

n_tg           <- nrow(tg)
n_green_laps   <- sum(lapstatus$FlagCondition != "Caution")
# true-green laps = settled green (8+ since restart); the honest denominator
# for the passing rate, since restart laps are a different regime.
n_truegreen_laps <- sum(lapstatus$FlagCondition == "Green (8+)")
pass_per_green <- round(n_tg / n_truegreen_laps, 2)

# ---- green-run segmentation ------------------------------------------------
# A green run = a maximal span of consecutive laps that are not stoppages.
green_laps <- lapstatus %>%
  filter(FlagCondition != "Caution") %>%
  arrange(Lap) %>%
  mutate(new_run = as.integer(c(1, diff(Lap)) != 1),
         run_id  = cumsum(new_run))

run_meta <- green_laps %>%
  group_by(run_id) %>%
  summarise(start = min(Lap), end = max(Lap), run_laps = n(), .groups = "drop") %>%
  mutate(run_label = sprintf("L%d–%d", start, end))

tg <- tg %>% left_join(green_laps %>% select(Lap, run_id), by = c("lap" = "Lap"))

# ---- incident reconstruction (multi-spot single-lap losses) ----------------
# Rebuild per-driver lap position from the pass-pairs (both roles carry
# prev/pos), compute lap-to-lap delta, flag big single-lap drops that are
# genuine on-track losses (not pit in/out laps, not stoppages).
pos_recs <- bind_rows(
  passes %>% transmute(lap, driver = passer, prev = passer_prev, pos = passer_pos),
  passes %>% transmute(lap, driver = passed, prev = passed_prev, pos = passed_pos)
) %>% distinct() %>% filter(!is.na(prev), !is.na(pos))

pit_set  <- pitlaps %>% distinct(driver, lap)
caut_set <- lapstatus %>% filter(FlagCondition == "Caution") %>% pull(Lap)

incidents <- pos_recs %>%
  mutate(delta = pos - prev) %>%                       # +ve = spots lost
  anti_join(pit_set, by = c("driver", "lap")) %>%
  anti_join(pit_set %>% mutate(lap = lap + 1), by = c("driver", "lap")) %>%
  filter(!lap %in% caut_set) %>%
  filter(delta >= 3)

n_incidents <- nrow(incidents)
```

<div class="hero">
<div class="hero-badge">GOLDEN TICKET SIMS</div>
<h1>Green-Flag Passing at the Brickyard</h1>
<p>Isolating true on-track green-flag passes — where the racing actually happened, when in the race, and who was moving forward on merit.</p>
</div>

## The headline numbers

<div class="kpi-row">
<div class="kpi"><div class="kpi-num">`r n_tg`</div><div class="kpi-lab">Green-flag passes</div></div>
<div class="kpi"><div class="kpi-num">`r pass_per_green`</div><div class="kpi-lab">Per green lap</div></div>
<div class="kpi"><div class="kpi-num">`r nrow(run_meta)`</div><div class="kpi-lab">Green runs</div></div>
<div class="kpi"><div class="kpi-num">`r n_green_laps`</div><div class="kpi-lab">Green laps</div></div>
<div class="kpi"><div class="kpi-num">`r n_incidents`</div><div class="kpi-lab">Multi-spot losses</div></div>
</div>

At roughly **`r pass_per_green` clean pass per green lap**, this was a track-position race — the leaders ran in clean-air trains while nearly all the on-track passing happened deep in the field.

## When did the passing happen?

```{r passing-by-lap, fig.height=5}
tg_by_lap <- tg %>% count(lap, name = "passes")
max_lap   <- max(lapstatus$Lap)

ggplot(tg_by_lap, aes(x = lap, y = passes)) +
  geom_col(fill = gts_green, width = 0.9) +
  scale_x_continuous(limits = c(0, max_lap + 1), expand = c(0.01, 0)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Green-flag passes by lap",
       subtitle = "Passing clusters early in each green run, then thins as the field strings out",
       x = "Lap", y = "Passes") +
  theme_gts()
```

## Passing per green run

```{r run-totals, fig.height=4.6}
run_totals <- tg %>%
  count(run_id, name = "passes") %>%
  right_join(run_meta, by = "run_id") %>%
  mutate(passes = coalesce(passes, 0L),
         per_lap = round(passes / run_laps, 2)) %>%
  arrange(start) %>%
  mutate(run_label = factor(run_label, levels = run_label))

ggplot(run_totals, aes(x = run_label, y = passes)) +
  geom_col(fill = gts_gold, width = 0.68) +
  geom_text(aes(label = sprintf("%d\n(%.2f/lap)", passes, per_lap)),
            vjust = -0.3, color = gts_text, fontface = "bold", size = 3.6, lineheight = 0.9) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Green-flag passes by run",
       subtitle = "Total passes in each green run, with per-lap rate",
       x = "Green run (laps)", y = "Passes") +
  theme_gts()
```

## Where on the leaderboard?

```{r by-position}
pos_band <- tg %>%
  mutate(band = cut(passer_pos,
                    breaks = c(0, 5, 10, 15, 20, 25, 40),
                    labels = c("1–5", "6–10", "11–15", "16–20", "21–25", "26+"))) %>%
  count(band)

ggplot(pos_band, aes(x = band, y = n, fill = n)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = n), vjust = -0.5, color = gts_text, fontface = "bold", size = 4) +
  scale_fill_gradient(low = "#2e5d34", high = gts_green, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Where passes happened on track",
       subtitle = "By the position the passer moved INTO",
       x = "Running position gained", y = "Passes") +
  theme_gts()
```

```{r position-distribution, fig.height=4.2}
ggplot(tg, aes(x = "", y = passer_pos)) +
  geom_violin(fill = gts_green, alpha = 0.35, color = gts_green, width = 0.8) +
  geom_boxplot(width = 0.16, fill = gts_panel, color = gts_gold, outlier.color = gts_amber) +
  scale_y_reverse(breaks = seq(0, 40, 5)) +
  coord_flip() +
  labs(title = "Spread of where passing happened",
       subtitle = "Distribution of running position after each pass — the mass sits deep in the field",
       x = NULL, y = "Position moved into") +
  theme_gts() +
  theme(axis.text.y = element_blank())
```

## Who was doing the passing?

```{r top-passers}
top_passers <- tg %>%
  count(passer, name = "passes_made") %>%
  arrange(desc(passes_made)) %>%
  slice_head(n = 15)

ggplot(top_passers, aes(x = reorder(passer, passes_made), y = passes_made)) +
  geom_col(fill = gts_gold, width = 0.72) +
  geom_text(aes(label = passes_made), hjust = -0.2, color = gts_text, fontface = "bold", size = 3.8) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_flip() +
  labs(title = "Most green-flag passes made",
       subtitle = "Top 15 · clean on-track passes only",
       x = NULL, y = "Passes made") +
  theme_gts()
```

```{r most-passed}
most_passed <- tg %>%
  count(passed, name = "times_passed") %>%
  arrange(desc(times_passed)) %>%
  slice_head(n = 15)

ggplot(most_passed, aes(x = reorder(passed, times_passed), y = times_passed)) +
  geom_col(fill = gts_red, width = 0.72) +
  geom_text(aes(label = times_passed), hjust = -0.2, color = gts_text, fontface = "bold", size = 3.8) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_flip() +
  labs(title = "Most often passed under green",
       subtitle = "Top 15 · drivers who lost the most spots on track",
       x = NULL, y = "Times passed") +
  theme_gts()
```

```{r net-passing}
made   <- tg %>% count(driver = passer, name = "made")
passed <- tg %>% count(driver = passed, name = "passed")
net <- full_join(made, passed, by = "driver") %>%
  mutate(across(c(made, passed), \(x) coalesce(x, 0L)),
         net = made - passed) %>%
  filter(made + passed >= 3) %>%
  arrange(desc(net))

net_plot <- bind_rows(slice_head(net, n = 10), slice_tail(net, n = 10)) %>%
  distinct() %>%
  mutate(dir = if_else(net >= 0, "Gained", "Lost"))

ggplot(net_plot, aes(x = reorder(driver, net), y = net, fill = dir)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = sprintf("%+d", net),
                hjust = if_else(net >= 0, -0.3, 1.3)),
            color = gts_text, fontface = "bold", size = 3.6) +
  scale_fill_manual(values = c(Gained = gts_green, Lost = gts_red), guide = "none") +
  geom_hline(yintercept = 0, color = gts_text, linewidth = 0.4) +
  coord_flip() +
  labs(title = "Net green-flag passing",
       subtitle = "Passes made minus times passed · on-track merit (min. 3 involvements)",
       x = NULL, y = "Net passes") +
  theme_gts()
```

## Multi-spot losses

Instances where a driver **lost 3 or more positions in a single green-flag lap** — not necessarily a mistake, but a sign of getting swamped, a moment lost, or a car falling off. `r n_incidents` happened across the race.

```{r incident-count}
inc_by_driver <- incidents %>%
  count(driver, name = "incidents") %>%
  arrange(desc(incidents)) %>%
  slice_head(n = 12)

ggplot(inc_by_driver, aes(x = reorder(driver, incidents), y = incidents)) +
  geom_col(fill = gts_amber, width = 0.7) +
  geom_text(aes(label = incidents), hjust = -0.25, color = gts_text, fontface = "bold", size = 3.8) +
  scale_y_continuous(breaks = pretty_breaks(), expand = expansion(mult = c(0, 0.14))) +
  coord_flip() +
  labs(title = "Who lost multiple spots in a lap most often",
       subtitle = "Single-lap losses of 3+ positions under green",
       x = NULL, y = "Occurrences") +
  theme_gts()
```

```{r incident-magnitude, fig.height=4.2}
ggplot(incidents, aes(x = factor(delta))) +
  geom_bar(fill = gts_red, width = 0.7) +
  geom_text(stat = "count", aes(label = after_stat(count)),
            vjust = -0.5, color = gts_text, fontface = "bold", size = 4) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.14))) +
  labs(title = "How bad were the single-lap losses?",
       subtitle = "Number of positions lost in one green lap",
       x = "Positions lost in a single lap", y = "Occurrences") +
  theme_gts()
```

```{r incident-table}
incidents %>%
  arrange(desc(delta)) %>%
  transmute(Lap = lap, Driver = driver,
            From = prev, To = pos, `Spots lost` = delta) %>%
  slice_head(n = 12)
```

<div class="footer">
Golden Ticket Sims · lap-by-lap green-flag pass analysis · Indianapolis Motor Speedway.
</div>
